---
title: Using terradactyl to Calculate Indicators of Rangeland Health and Conditions
  in R
author: "Joe Brehm"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview
In this session we will walk through the process of using terradactyl to calculate rangeland indicators. There are two steps in this process, the gather stage and the calculation stage. The gather stage reformats raw data into a "tall" format, the calculation step uses that tall data to create the indicators.

- Overall input: sample from terradat and species list
- Overall output: tall tables and indicator data
- Who would need to use the gather stage, who could skip to the calculation stage
- Use LPI as an example (maybe. It is also the most complicated one.)

?? How much to explain "tall" data

?? Who talks about PrimaryKey

?? Should we include species list QC? 


## 1: Gather functions

Gather functions are step 1 of the process, beginning with raw AIM or LMF format data and ending with harmonized data in a tall format, ready for use with terradactyl. 

- Input: sample from terradat
- Output: tall tables
- Discuss AIM & LMF style input
- Demonstrate with direct link to GDB and with tables loaded into R
- Use gap and Header as examples, refer to the others

!! Make the data subset.

!! Make sure people understand how to set the paths up on their machines

```{r paths}
# path_gdb <- "C:\Users\jrbrehm\Documents\terradactyl_demo_SRM2024\rawdata.gdb"
# path_tall <- "C:\Users\jrbrehm\Documents\terradactyl_demo_SRM2024\tall"
# path_indicators <- "C:\Users\jrbrehm\Documents\terradactyl_demo_SRM2024\indicators"
# path_specieslist <- "C:\Users\jrbrehm\Documents\terradactyl_demo_SRM2024\demo_species_list.csv"
# 
# library(terradactyl)
# library(sf)
```

- Load in data from geodatabase
- demonstrate that it looks different and needs harmonizing
- discuss where people can get their own tblPlots / PINTERCEPT data.

``` {r}
# tblPlots <- sf::st_read(dsn = path_gdb, layer = "tblPlots") # will produce ignorable warning
# POINT <- sf::st_read(dsn = path_gdb, layer = "POINT") # will produce ignorable warning
# 
# tblPlots[x:y]
# POINT[x:y]
```

### 1a: Gathering header data

- Gather header data
- demonstrate that it is now compatible

```{r}
# header_1 <- gather_header(tblPlots = tblPlots, source = "AIM")
# header_2 <- gather_header(POINT = POINT, source = "LMF")
# head(header_1)
# head(header_2)
```

- demonstrate that the gather functions can be sent directly to a gdb

```{r}
# header_3 <- gather_header(dsn = path_gdb, source = "AIM")
# header_4 <- gather_header(dsn = path_gdb, source = "LMF")
# 
# (header_1 == header_3) & (header_2 == header_4)
```


### 1b: Gathering gap data
- load gap data
- why raw data is hard to work with #1: data in multiple columns

!! list all the other methods in prose or code? Do this elsewhere?

!! add LMF?

``` {r}
# tblLPIHeader <- sf::st_read(dsn = path_gdb, layer = "tblLPIHeader")
# tblLPIDetail <- sf::st_read(dsn = path_gdb, layer = "tblLPIDetail")
# 
# sample_pk <- tblLPIHeader$PrimaryKey[1]
# 
# tblLPIDetail %>% dplyr::select("lower1", "lower2", etc) %>% dplyr::filter(PrimaryKey == sample_pk)
```

- why raw data is hard to work with #2: data in multiple tables

```{r}
# colnames(tblLPIHeader)
```

- gather gap
- save output

```{r}
# lpi_tall <- gather_lpi(tblLPIHeader = tblLPIHeader, tblLPIDetail = tblLPIDetail, source = "AIM")
# saveRDS(lpi_tall, file.path(path_tall, "lpi_tall.rdata"))
# saveRDS(header1, file.path(path_tall, "header.rdata"))
```

- Talk about the structure of the tall data

``` {r}
# colnames(lpi_tall)
# nrow(lpi_tall)
# length(unique(lpi_tall$PrimaryKey))
# lpi_tall[,c("PrimaryKey", "code", "layer")] # add some columns for useful header data too
```

## 2: Indicator functions
Indicator functions are step 2 of the terradactyl workflow, beginning with the harmonized tall data produced by the gather functions and ending with analysis-ready indicator data.

- Input: tall tables
- Output: indicators
- Go through gap functions

?? build_indicators first as a "start here" or last as a capstone?

?? How much to go into lpi specifics? None at all, almost certainly. See how you are doing on time. 

!! pct_cover? If any, then this

!! custom indicators? Maybe, would probably require species_join

!! first vs any? No

!! tall vs wide? Maybe

!! by_line? No

!! Other cover functions (cover_between_plant, pct_cover_species, etc)? Mention briefly but stick to general case of pct_cover


```{r}
# lpi_indicators <- lpi_calc(header = file.path(path_tall, "header.rdata"),
#                            lpi_tall = file.path(path_tall, "lpi_tall.rdata"),
#                            species_file = path_specieslist,
#                            source = "AIM",
#                            dsn = NULL)### WHY IS DSN REFERENCED HERE? IT SHOULD NOT BE. CHECK TERRADACTYL
# 
# colnames(lpi_indicators)
# lpi_indicators$pickacol
```
