---
title: "R in Rangelands Climate Engine"
output: html_notebook
---

```{r}
library(tidyverse) # Data processing and visualization
library(httr) # HTTP API requests
library(httr2) # HTTP API requests
library(DT) # Formatting data tables
library(sf) # Importing shapefile
library(mapview) # Visualizing shapefile
library(geojsonsf)
```

## Authenticate to API and run preliminary test

Define Climate Engine API key and run simple endpoint to get key expiration to ensure that the key authenticated correctly:

```{r}
# Define root url for Climate Engine API
root_url <- "https://api.climateengine.org/"

# Define key
key = readLines('ce_key.txt')

# Define endpoint
endpoint = '/home/key_expiration'

# Run simple endpoint to get key expiration
test_r <- GET(paste0(root_url, endpoint), config = add_headers(Authorization = key))
print(content(test_r, "text",  encoding = "ISO-8859-1"))

# Clean up unneeded objects
rm(test_r, endpoint)
```

## Import the wind erosion area of interest

This is a sample AOI in southern New Mexico, but you can run similar analysis for any shapefile. We will import the shapefile as an sf object (and upload it to Google Earth Engine) and visualize them on a leaflet map using the mapview package. You can also modify which area of interest you want to analyze in this code chunk.

```{r}
# Import the field boundaries shapefile and select an individual area of interest. 
aoi <- st_read('MapLayers/SandyESG.shp') %>%
  slice(4) %>%
  st_simplify(preserveTopology = TRUE, dTolerance = 10) %>% 
  st_transform(crs = "OGC:CRS84") 

aoi_name = aoi$areasymbol %>% unlist()

# Visualize the allotments on a leaflet map using mapview package
mapview(aoi)
```

## Make API requests to get Rangeland Analysis Platform (RAP) fractional vegetation cover data for annual forbs and grasses (AFG), perennial forbs and grasses (PFG), shrubs (SHR), trees

Here we will iterate over a list of vegetation types to get an annual timeseries of vegetation cover for our area of interest. There are a few helper functions included here to clean up the data returned from the API. Requests are sent to the API in the JSON format and the resulting data is returned in the JSON format, so the helper functions translate the JSON into an R dataframe.

```{r}
# Define dataset for function below
# Documentation on the Rangeland Analysis Platform is available through rangelands.app
# API Documentation on the Rangeland Analysis Platform: https://docs.climateengine.org/docs/build/html/variables.html#rst-rap-vegetation-cover
dataset = "RAP_COVER"

# Make a list of ET Model options — we will loop over these using the function below to return all of the data
variables <- c('AFG', 'PFG', 'SHR', 'TRE', 'BGR')

# Function to return data frame of OpenET time-series
get_rap_df <- function(variable){
  
  print(paste0("Requesting data for ", variable))
  
  # Define Climate Engine API Point 
  # Documentation for this endpoint found here: https://docs.climateengine.org/docs/build/html/timeseries.html#rst-timeseries-native-custom-asset
  endpoint <- 'timeseries/native/custom_asset'
  
  # Create list of API parameters for request
  params = list(
    dataset = dataset,
    variable = variable,
    start_date = '2000-01-01',
    end_date = '2022-12-31',
    area_reducer = 'mean',
    asset_id = 'users/EricRJensen-DRI/SandyESG',
    sub_choices = paste0('["', aoi_name, '"]'),
    filter_by = 'areasymbol')

  # Make API request
  data_json <- GET(paste0(root_url, endpoint), config = add_headers(Authorization = key), query = params)

  # Parse the JSON returned from API to prepare dataframe
  data_json_parse <- content(data_json, "parsed")
  data_df <- enframe(unlist(data_json_parse[[1]]$Data))
  data_cols <- data_df$name %>% unique()

  # Function to iterate over "name" values in parsed dataframe and return single column dataframe with those values
  generate_data_frame <- function(col, df){
  
    # Filter for name from parsed API return
    df_val <- df %>%
      filter(name == col)
  
    # Create output data frame based on name from parsed API return
    out_df <- tibble(val = df_val$value)
  
    # Provide names from parsed API return as column name
    colnames(out_df) <- c(col)
  
    return(out_df)
  }

  # Map generate_data_frame over list of columns and clean up resulting data frame
  variable_r_df <- map(data_cols, generate_data_frame, data_df) %>%
    bind_cols() %>%
    mutate(Date = as.Date(Date),
           Value = as.numeric(.[[2]]),
           Variable = paste0(variable)) %>%
    filter(Value > -10) %>%
    select(-2)
  
  return(variable_r_df)
}

# Map OpenET API function over list of models to get timeseries from all models
rap_df <- map(variables, get_rap_df) %>%
  bind_rows() # Combine list of dataframes into an individual dataframe
datatable(rap_df)
```

## Make API requests to get Rangeland Analysis Platform (RAP) fractional vegetation cover data for annual forbs and grasses (AFG), perennial forbs and grasses (PFG), shrubs (SHR), trees

Make simple plots of the RAP time-series data, similar to the RAP web application

```{r}
# Make a simple plot of the time-series data
ggplot(data = rap_df)+
  geom_line(mapping = aes(x = Date, y = Value, color = Variable))+
  labs(title = paste0('RAP Cover for ', aoi_name), y = 'Fractional cover (%)') +
  scale_colour_manual(values = c('AFG' = "#F8766D", 'BGR' = "#B79F00", 'PFG' = "#00BA38", 'TRE' = "#00BFC4", 'SHR' = "#619CFF"), labels = c("AFG", "BGR", "PFG", "TRE", "SHR"))+
  theme_minimal()+
  theme(legend.position="bottom",
        axis.title.x = element_blank(),
        legend.title=element_blank())
```


## Get 180-day Standardized Precipitation-Evapotranspiration Index (SPEI) data from gridMET Drought.

Here, we will get histograms of 180-day SPEI drought data based on accepted drought categories developed by NOAA NIDIS (drought.gov). This will allow us to make a beautiful stacked bar chart to depict the drought history for our AOI. You can learn more about drought indicators and how the Climate Engine team is helping the BLM to assess drought on our support site: https://support.climateengine.org/article/128-rangeland-drought-assessment

```{r}
# Define Climate Engine API Point 
# Documentation for this endpoint found here: https://docs.climateengine.org/docs/build/html/zonal_statistics.html#rst-zonal-stats-pixel-count-custom-asset
endpoint <- 'zonal_stats/pixel_count/custom_asset'

# Documentation on the gridMET Drought is available here: https://support.climateengine.org/article/45-gridmet-drought
# API Documentation on the Rangeland Analysis Platform: https://docs.climateengine.org/docs/build/html/variables.html#rst-gridmet-drought
dataset <- 'GRIDMET_DROUGHT'
variable <- 'spei180d'

# Get date list to iterate over
yearlist = seq(from = 2000, to = 2004, by = 1)

generate_date_seq <- function(year){
  date_seq_year <- seq(from = as.Date(paste0(as.character(year), "/1/5")), to = as.Date(paste0(as.character(year), "/12/31")), by = '5 days')
  return(c(as.character(date_seq_year)))
}

date_seq <- map(yearlist, generate_date_seq)
date_seq %>% flatten_chr()

# Generate dataframe of histograms
all_df <- tibble()
get_drought_histogram <- function(date_in){
  
  print(paste0('Running ', date_in))
        
  # Create list of API parameters for request
  params = list(
    dataset = dataset,
    variable = variable,
    end_date = date_in,
    area_reducer = 'mean',
    asset_id = 'users/EricRJensen-DRI/SandyESG',
    sub_choices = paste0('["', aoi_name, '"]'),
    filter_by = 'areasymbol')
  
  # Make API request
  variable_r <- GET(paste0(root_url, endpoint), config = add_headers(Authorization = key), query = params)
  
  # Parse JSON returned from API
  variable_r_parse <- content(variable_r, "parsed")
  
  # Generate dataframe of histogram for date
  date_out = variable_r_parse[[1]]$Date
  print(date_out)
  bin_length = length(variable_r_parse[[1]]$spei180d)
  print(bin_length)
  row_df <- tibble('Date' = date_out)
  print(row_df)
  for (i in seq(1, bin_length, 1)){
    print('i',i)
    bin = as.character(i)
    val = variable_r_parse[[1]]$spei180d[i][[1]]
    df = tibble(bin = val)
    names(df) <- bin
    row_df = bind_cols(row_df, df)}
    
  print(row_df)
}

get_drought_histogram(date_seq[1])

```



