---
title: "R in Rangelands Wkshp"
author: "Georgia Harrison, Leah Dreesmann, Claire Tortorelli"
date: '2024-01-31'
output:
  pdf_document: default
  html_document: default
---


# Welcome to Putting the R in Rangelands!
### Access the data and interactive code via [github](https://github.com/Landscape-Data-Commons/RinRangelands)


###################################################################################
# Part 1: Getting started with R

# Before you begin:
## Install R and RStudio
As of January 3, 2024, the current R version is 4.3.2 and the current RStudio version is 2023.12.0+369.
You should download current versions of both these programs and install them on your computer if you haven’t already. You can download R from CRAN and the free version of RStudio from their site.


# Why use R? 
* Free
* open source, always growing an evolving
* tons of packages
* great for data analysis, statistics and data visualization 

# R vs RStudio? 
* **R** is the software that performs instructions. 
* **RStudio** is an interface to interact with R.


## RStudio
Let's take a tour through R studio
* **Console** pane - contains the output
    push enter to run 
    for example, run 4 + 4
      output is in brackets on the next line [8]
* **Source** pane - for working with scripts you want to save
    All files end in .R
    Run section using RUN button, or Ctrl+Enter (or Cmd+Enter for Macs)
* Also take  look at the environment & files panes. 
  

# Quick terminology
* **Package** base R is bare bones. Packages are like apps that other people have made that you can download and use. Some are standard for most R users, others are subject specific
* **Directory**  The working directory is your home base for this R session.
* **Environment** your current workspace. This includes any files you have read into R, and data tables you created, etc


# What is up with all the hashtags
```{r}
## these are comments and notes
# anything following a hashtag is a comment. This is just text
# use this for notes within code, troubleshooting 
```


# R scripts vs R Markdown (RMD)? 
* In regular **R scripts**, outputs are outside of the script: in the console, plots or other windows.  These files end with .R
* In **R Markdown**, the script is broken into text and code sections, called chunks. You can format the text (hence all the asterisks), and outputs are spit out under code chunks. To run a chunk of code, hit the green triangle in each chunk, or run all chunks using the upper Run arrow. These files end with .Rmd


# The Tidyverse
This workshop will be focused on using tools within the  [tidyverse](https://www.tidyverse.org/) for data manipulation and visualization. The tidyverse is a collection of packages that all play nice with eachother - they have the same philosophy, grammar, and data structures. This allows for fast, efficient workflows in R. 
The world of R is moving to the tidyverse, and you should too! 

## Packages within the tidyverse: 
* Importing data (readr)
* Data manipulation (dplyr, tidyr)
* Working with data types (stringr for strings, lubridate for date/datetime, forcats for categorical/factors)
* Data visualization (ggplot2)
* Data-oriented programming (purrr)
* Communication (Rmarkdown, shiny)



# Install packages
There are a few ways you can install a package:

1) In packages tab, using the install button. Then library a package (aka activate it for that session) by checking the box
2) Using install.packages("packagename")
```{r}
# install.packages("tidyverse") #run this! only once, good practice to run in the console
```
**note, only do this once!** 
After installation, library a package to recall it and have it ready to use during that session
```{r, message=FALSE, warning = FALSE}
library(tidyverse)
```
### the tidyverse is special because it is actually a family of packages, so this function is actually calling many packages

# Help
Here are a few of the key ways to get help in R. 
1. **Help function**: to pull up the documentation for a package or function:
?package
?functionname

for example: 
```{r}
?readr
?mutate
```
These bring you to **R documentation**, which is always set up in the same way with these sections:
* Description 
* Usage
* Arguments
* Examples or Details


2. **Stack overflow**: R questions and answers
in most cases, someone else has already asked your question 

There are [cheatsheets](https://posit.co/resources/cheatsheets/) for each tidyverse package (and many other R entities!)



# Working Directory
This is where R, by default, will go to look for any datasets you load and is the place R will save files you save. When working on a simple project, I save my R scripts and all files related to that project into a single folder that I set as my working directory. This makes it so I don’t have to write out the whole directory path every time I want to load or save something. This also helps me keep organized.

You can check the current working directory: 
```{r}
getwd()
```
You can do this with code of with the control panels: 
Session > set working directory > to source file location >  navigate to your home base folder on your computer
OR 
Files pane > More.. > Set working directory

# Let's all set the working directory for this workshop 
Important: You must either use single forward slashes or double backslashes in the directory path in R instead of the single backslashes. If you work in Windows this will not be what you are used to.
setwd("C:/Users/Aosmith/R workshops/r-basics-workshop") OR
setwd("C:\\Users\\Aosmith\\R workshops\\r-basics-workshop")
```{r}
setwd("C:\\Users\\harr4718\\OneDrive - University of Idaho\\R code (copies)\\SRM R workshop\\DemoData")
#customize to your own file path
```



# Read in the data
let's get started actually working with the data 
First step is to import the data using readr (a package within the tidyverse) 
using the *read_csv()* function 
These data were obtained from the Landscape Data Commons, which we will talk about more in Part 2
```{r}
starter_data = read_csv("Part1/RinRange_Pt1Data.csv")

```



###################################################################################
\newpage
